\section{Canonical Form and the \texorpdfstring{$\kappa \leftrightarrow 1/\kappa$}{kappa <-> 1/kappa} Symmetry}
\label{sec:canonical}

\subsection{The Change of Variables}

Fisher's paired efficiency \parencite{fisher1935} is the benchmark case $\kappa = 1$; when the two margins differ, the ratio-of-variances literature \parencite{bose1935} and the Pitman--Morgan test for equal paired variances \parencite{pitman1939,morgan1939} address related distributional questions but not the canonical efficiency coordinate developed below.

The Paired Efficiency Factor \parencite{brownPEFempirical} is defined as the ratio of unpaired to paired variance,
\begin{equation}
  \eta(\kappa,\rho)
  = \frac{1+\kappa}{1+\kappa-2\sqrt{\kappa}\,\rho},
  \label{eq:pef_original}
\end{equation}
where $\kappa = \sigma_{\mathrm{B}}^{2}/\sigma_{\mathrm{A}}^{2} > 0$ is the variance ratio and $\rho \in (-1,1)$ is the Pearson correlation.  Setting
\begin{equation}
  \tau = \tfrac{1}{2}\log\kappa,
  \label{eq:tau_def}
\end{equation}
so that $\sqrt{\kappa} = e^{\tau}$ and $1/\sqrt{\kappa} = e^{-\tau}$, the denominator of \cref{eq:pef_original} becomes
\[
  1 + \kappa - 2\sqrt{\kappa}\,\rho
  = e^{-\tau}(e^{\tau} + e^{-\tau}) e^{\tau} - 2\rho e^{\tau} e^{-\tau} \cdot \tfrac{e^{\tau}+e^{-\tau}}{e^{\tau}+e^{-\tau}}
\]
and, after dividing numerator and denominator by $\sqrt{\kappa} = e^{\tau}$, we obtain the \emph{canonical form}:
\begin{equation}
  \eta(\tau,\rho)
  = \frac{\cosh\tau}{\cosh\tau - \rho}.
  \label{eq:canonical}
\end{equation}
Here $\tau \in \mathbb{R}$ parameterises variance asymmetry on a symmetric log-scale, whilst $\rho$ retains its role as correlation.  The half-strip $\{\tau \ge 0,\, \rho \in (-1,1)\}$ is the \emph{fundamental domain}: the original parameterisation $(\kappa,\rho)$ with $\kappa \ge 1$ maps bijectively onto it.  Everything in Sections~\ref{sec:mobius}--\ref{sec:fisher_rao} is a consequence of this identity.

\subsection{The \texorpdfstring{$\kappa \leftrightarrow 1/\kappa$}{kappa <-> 1/kappa} Involution}

\begin{theorem}[Involution]
\label{thm:involution}
For all $\kappa > 0$ and $\rho \in (-1,1)$ with $\cosh\tau \ne \rho$,
\[
  \eta(\kappa,\rho) = \eta(1/\kappa,\rho).
\]
\end{theorem}

\begin{proof}
Substituting $\kappa \mapsto 1/\kappa$ replaces $\tau = \tfrac{1}{2}\log\kappa$ by $-\tau$.  Since $\cosh$ is even, $\cosh(-\tau)=\cosh\tau$, and the right-hand side of \cref{eq:canonical} is unchanged.
\end{proof}

\begin{corollary}
The map $\kappa \mapsto 1/\kappa$ is an involution on the parameter space that acts trivially on $\eta$.  The four-quadrant taxonomy of \textcite{brownPEFempirical} partitions $(\kappa,\rho)$ by the signs of $(\kappa-1)$ and $\rho$; the involution exchanges Quadrants~1 and~2, and Quadrants~3 and~4, whilst fixing $\eta$.  Consequently, the four-quadrant structure is a \emph{two-to-one covering} of the fundamental domain $\tau \ge 0$; all geometric content resides in the half-strip.
\end{corollary}

The involution is validated empirically in \cref{sec:numerical}: across all 113 sports KPI studies and 19\,903 non-sports units, the estimator-level residual $|\hat{\eta}(\kappa,\rho) - \hat{\eta}(1/\kappa,\rho)|$ is at most $1.78 \times 10^{-15}$, consistent with machine-precision equality.

\subsection{Domain and Singularity Locus}

The denominator $\cosh\tau - \rho$ vanishes when $\rho = \cosh\tau \ge 1$, which is outside the physical domain $\rho \in (-1,1)$ except at the single boundary point $(\tau,\rho) = (0,1)$, where $\eta \to +\infty$.  Near this pole, set $\tau = \varepsilon_1$ and $\rho = 1 - \varepsilon_2$ with $\varepsilon_1, \varepsilon_2 \to 0^+$.  Then $\cosh\tau - \rho \approx (1 + \tfrac{1}{2}\varepsilon_1^2) - (1-\varepsilon_2) = \varepsilon_2 + \tfrac{1}{2}\varepsilon_1^2$, giving the local parabolic blow-up
\begin{equation}
  \eta \;\approx\; \frac{1}{(1-\rho) + \tfrac{1}{2}\tau^{2}}.
  \label{eq:parabolic_blowup}
\end{equation}
Iso-$\eta$ contours near the pole are therefore parabolas $1-\rho = C - \tfrac{1}{2}\tau^2$ in the $(\tau,\rho)$-plane, not circles.  This geometric fact is made precise in \cref{sec:sphere}, where the pole is identified with a marked point on the Riemann sphere.

\subsection{Differential Structure}

From \cref{eq:canonical},
\begin{align}
  \frac{\partial\eta}{\partial\rho}
  &= \frac{\cosh\tau}{(\cosh\tau-\rho)^{2}} > 0,
  \label{eq:deta_drho}\\[4pt]
  \frac{\partial\eta}{\partial\tau}
  &= \frac{-\sinh\tau\,(\cosh\tau-\rho) + \cosh\tau\sinh\tau}{(\cosh\tau-\rho)^{2}}
   = \frac{\rho\sinh\tau}{(\cosh\tau-\rho)^{2}}.
  \label{eq:deta_dtau}
\end{align}
Thus $\eta$ is strictly increasing in $\rho$ for all $\tau$.  In $\tau$, the sign of $\partial\eta/\partial\tau$ matches the sign of $\rho$: $\eta$ increases with $|\tau|$ when $\rho > 0$ and decreases when $\rho < 0$.  The critical set $\partial\eta/\partial\tau = 0$ is the union of the lines $\tau = 0$ and $\rho = 0$.  There are no interior critical points in $(\tau,\rho) \in (0,\infty)\times(-1,0)\cup(0,1)$.  Sensitivity to correlation, measured by \cref{eq:deta_drho}, is maximised at $\tau = 0$ (Fisher's equal-variance case) and decays monotonically as $|\tau| \to \infty$.

\subsection{Special Values}
\label{sec:special_values}

\begin{table}[t]
  \centering
  \caption{Boundary behaviour of the canonical form~\cref{eq:canonical}.}
  \label{tab:special_values}
  \begin{tabular}{@{}llll@{}}
    \toprule
    \textbf{Condition} & \textbf{$\tau$} & $\eta$ & \textbf{Note} \\
    \midrule
    $\kappa = 1$ (Fisher case) & $0$ & $\dfrac{1}{1-\rho}$ & Classical formula \\[6pt]
    $\rho = 0$                 & any & $1$                   & Correlation-free baseline \\[2pt]
    $\rho \to +1$              & $0$ & $+\infty$             & Perfect pairing, equal variances \\[2pt]
    $\rho \to -1$              & $0$ & $\tfrac{1}{2}$        & Maximum variance amplification \\[2pt]
    $\kappa \to 0$ or $\infty$ & $|\tau| \to \infty$ & $1$ & Extreme asymmetry collapses to baseline \\
    \bottomrule
  \end{tabular}
\end{table}

A notable identity follows from evaluating \cref{eq:canonical} at $\rho = +1$ and $\rho = -1$:
\begin{equation}
  \eta(\tau,+1)\cdot\eta(\tau,-1)
  = \frac{\cosh\tau}{\cosh\tau-1}\cdot\frac{\cosh\tau}{\cosh\tau+1}
  = \frac{\cosh^{2}\tau}{\cosh^{2}\tau-1}
  = \frac{\cosh^{2}\tau}{\sinh^{2}\tau}
  = \coth^{2}\tau.
  \label{eq:reciprocal_product}
\end{equation}
At $\tau = 0$ (Fisher's case), $\coth\tau \to \infty$, reflecting the pole.  For large $|\tau|$, $\coth\tau \to 1$, so the product approaches unity: the efficiency gain at maximal positive correlation is asymptotically the reciprocal of the efficiency loss at maximal negative correlation.  This is a residual echo of the involution.

\subsection{Consequence for the Empirical Four-Quadrant Taxonomy}
\label{sec:quadrant_coverage}

The empirical paper partitions the $(\kappa,\rho)$ plane by the signs of $\kappa - 1$ and $\rho$, yielding four quadrants.  \Cref{thm:involution} shows this partition is a coordinate artefact: the pairs (Q1, Q2) and (Q3, Q4) each collapse to a single region under the substitution $\kappa \mapsto \max(\kappa,1/\kappa)$.  Equivalently, viewing the parameter space through the canonical coordinates $(\tau,\rho)$ with $\tau \ge 0$, the four-quadrant taxonomy reduces to the two regimes $\rho > 0$ and $\rho < 0$.  The qualitative difference between the regimes -- explored in the partition-function context in \cref{sec:partition} and in the Fisher--Rao context in \cref{sec:fisher_rao} -- is therefore a genuine feature of the geometry, whilst the Q1-vs-Q2 and Q3-vs-Q4 distinctions are artefacts of the original $(\kappa,\rho)$ chart.

This has a concrete empirical implication: pooling estimators $\hat{\kappa} \mapsto \max(\hat{\kappa},1/\hat{\kappa})$ across the six datasets of \textcite{brownPEFempirical} must collapse Q1~$\leftrightarrow$~Q2 and Q3~$\leftrightarrow$~Q4 within bootstrap tolerance.  The symmetry audit (\cref{sec:numerical}) validates this at machine precision.
