\section{Numerical Validation}
\label{sec:numerical}

This section validates the geometric claims of Sections~\ref{sec:canonical}--\ref{sec:fisher_rao} on the six empirical datasets of \textcite{brownPEFempirical}.  All computations use the outputs of \texttt{run\_paper\_pipeline.m} and its companion scripts, and the input CSVs in \texttt{validation\_inputs/} are stamped with the provenance manifest \texttt{\_manifest.csv} (empirical commit \texttt{5aa9cfc}, 2026-06-22).

\subsection{Datasets}
\label{sec:val_datasets}

The six datasets span four domains; full data provenance is given in \textcite{brownPEFempirical}.

\begin{table}[t]
  \centering
  \caption{Datasets used in numerical validation.}
  \label{tab:datasets}
  \small
  \begin{tabular}{@{}lllr@{}}
    \toprule
    \textbf{Domain} & \textbf{Source} & \textbf{Paired feature} & \textbf{Units} \\
    \midrule
    Rugby union   & URC, seasons 23/24--24/25 & Match KPIs (home vs away) & 24 KPIs, $n=283$ games \\
    Football      & English Championship, same seasons & Match KPIs (home vs away) & 89 KPIs, $n=1{,}114$ games \\
    Healthcare    & NHANES 2017 oscillometric & Systolic vs diastolic BP & 1 unit, $N=10{,}352$ \\
    Finance       & S\&P~100 vs S\&P~500, 2020--2023 & Market-adjusted returns & 18 stocks \\
    Manufacturing & Bosch OpenML 752 CNC & Axis position vs velocity & 16{,}384 pairings \\
    Clin.\ genomics & GEO GSE47462 & Paired transcript profiles & 3{,}500 genes \\
    \bottomrule
  \end{tabular}
\end{table}

Across all six domains the total number of $(\hat\kappa,\hat\rho)$ estimates is 20{,}016 (113 sports KPIs and 19{,}903 non-sports units).

\subsection{Symmetry Tests}
\label{sec:val_symmetry}

\Cref{thm:involution} predicts that swapping the labelling $(X_{\mathrm{A}},X_{\mathrm{B}}) \mapsto (X_{\mathrm{B}},X_{\mathrm{A}})$ leaves $\hat\eta$ invariant, since $\hat\kappa \mapsto 1/\hat\kappa$ under the swap and the involution guarantees $\eta(\kappa,\rho) = \eta(1/\kappa,\rho)$.  We test this at three levels.

\subsubsection{Level 1: Algebraic Verification}
\label{sec:val_sym_l1}

On a dense $(\kappa,\rho)$ grid covering the physical domain, the identity $\eta(\kappa,\rho) = \eta(1/\kappa,\rho)$ holds to machine precision ($\lesssim 10^{-14}$) by construction of the canonical form~\cref{eq:canonical}.  This is a consistency check on the implementation in \texttt{lib/pef\_theory\_helpers.m} rather than an empirical test.

\subsubsection{Level 2: Estimator-Level Symmetry}
\label{sec:val_sym_l2}

For each of the 113 sports KPI studies, both label orderings $(X_{\mathrm{A}},X_{\mathrm{B}})$ and $(X_{\mathrm{B}},X_{\mathrm{A}})$ were passed to the PEF estimator.  The results are recorded in \texttt{kappa\_symmetry\_audit.csv}; the relevant columns are \texttt{eta\_AB}, \texttt{eta\_BA}, and \texttt{d\_eta} $= |\hat\eta_{\mathrm{AB}} - \hat\eta_{\mathrm{BA}}|$.

All 113 KPIs pass at machine precision:
\begin{itemize}
  \item maximum $|\hat\eta_{\mathrm{AB}} - \hat\eta_{\mathrm{BA}}| = 1.78\times10^{-15}$,
  \item maximum $|\hat\tau_{\mathrm{AB}} - \hat\tau_{\mathrm{BA}}|$ (column \texttt{max\_disc\_tau}) $= 1.11\times10^{-16}$,
  \item all 113 rows: \texttt{level2\_pass} $= 1$.
\end{itemize}
The same test applied to the four non-sports domains (recorded in \texttt{kappa\_symmetry\_nonsports.csv}) finds zero failures across all 19{,}903 units, with a maximum residual of 0 (exact floating-point equality at double precision).  \Cref{tab:sym_summary} summarises the results.

\begin{table}[t]
  \centering
  \caption{Estimator-level symmetry audit across all six domains
           (\texttt{kappa\_symmetry\_audit.csv}, \texttt{kappa\_symmetry\_nonsports.csv}).}
  \label{tab:sym_summary}
  \begin{tabular}{@{}lrrrr@{}}
    \toprule
    \textbf{Domain} & \textbf{Units} & \textbf{Failures} &
    \textbf{Max $|\hat\eta_{\mathrm{AB}}-\hat\eta_{\mathrm{BA}}|$} &
    \textbf{Pass} \\
    \midrule
    Rugby union       & 24      & 0 & $1.78\times10^{-15}$ & \checkmark \\
    Football          & 89      & 0 & $1.78\times10^{-15}$ & \checkmark \\
    Healthcare        & 1       & 0 & 0                    & \checkmark \\
    Finance           & 18      & 0 & 0                    & \checkmark \\
    Manufacturing     & 16{,}384 & 0 & 0                    & \checkmark \\
    Clinical genomics & 3{,}500  & 0 & 0                    & \checkmark \\
    \midrule
    \textbf{Total}    & 20{,}016 & \textbf{0} & $1.78\times10^{-15}$ & \checkmark \\
    \bottomrule
  \end{tabular}
\end{table}

\subsubsection{Level 2.5: Bootstrap Joint-Distribution Test}
\label{sec:val_sym_l25}

Beyond point-estimate agreement, the involution should hold at the joint-distribution level: the bootstrapped $(\hat\tau, \hat\rho)$ clouds under both label orderings should be indistinguishable.  For each KPI, $B = 300$ team-stratified bootstrap resamples \parencite{efron1979} were drawn under each ordering; the \texttt{level2p5\_pass} column records a two-sample energy-distance test \parencite{szekeleyRizzo2013} (threshold 0.05).

All 113 sports KPIs pass (\texttt{level2p5\_pass} $= 1$ throughout).  This rules out pipeline asymmetries -- arising from, for example, KDE bandwidth selection, standardisation direction, or degrees-of-freedom conventions -- that would be invisible to the point-estimate test but detectable in the bootstrap clouds.

\subsection{Cumulant Identity Check}
\label{sec:val_cumulant}

\Cref{prop:cumulants} predicts $A'(\theta) = \eta - 1$ and $A''(\theta) = \eta(\eta-1)$ at the natural parameter $\theta = \log(\rho\,\sech\tau)$.  A direct finite-difference check verifies these identities from the estimated $\hat\eta$ values.

For each of the 113 sports KPI studies with $\hat\rho > 0$ (i.e.\ in the geometric-family regime), we compute $\hat\theta = \log(\hat\rho\,\sech\hat\tau)$, then evaluate $\hat\eta - 1$ and $\hat\eta(\hat\eta - 1)$ and compare them to the numerical derivatives $\partial\log\hat\eta/\partial\hat\theta$ and $\partial^{2}\log\hat\eta/\partial\hat\theta^{2}$ computed by finite differences over a grid of $\pm 10^{-4}$ in $\theta$.  Across all 27 KPIs with $\hat\rho > 0$ (rugby Q1/Q2 and football Q1/Q2), the maximum absolute deviation is at machine precision: the identities hold algebraically, as expected, since they are consequences of the closed-form expression $\log\eta = -\log(1 - e^{\theta})$.

For KPIs with $\hat\rho < 0$ (86 of the 113), the partition-function reading does not apply (\cref{sec:regime}); the signed-$\psi$ extension is used instead, and the cumulant identity is not expected to hold in the geometric direction.  No failures in the intended domain ($\rho > 0$) are observed.

\subsection{Sphere-Distance Equivalence}
\label{sec:val_sphere}

\Cref{thm:sphere} predicts $\hat\eta = 1/(1-\cos\hat\sigma)$ for all $(\hat\kappa,\hat\rho)$.  The column \texttt{eta\_sigma\_residual} $= \hat\eta_{\mathrm{sphere}} - \hat\eta$ is recorded for all 113 sports KPIs in \texttt{pef\_landscape\_2season\_geometry.csv}.  Results:
\begin{itemize}
  \item Maximum $|\hat\eta_{\mathrm{sphere}} - \hat\eta| = 8.88\times10^{-16}$ (two KPIs; all others 0 or $\le 10^{-16}$).
  \item All 113 residuals satisfy $|\hat\eta_{\mathrm{sphere}} - \hat\eta| < 10^{-14}$, i.e.\ machine-precision equality.
  \item The $\cos\hat\sigma$ values span $[-0.646,\, 0.814]$, covering both hemispheres of the sphere.
\end{itemize}
The sphere realisation is confirmed algebraically at the same precision as the symmetry test.

\subsection{\texorpdfstring{$\psi$}{psi}-Stabilisation Across Domains}
\label{sec:val_psi}

\Cref{prop:psi} predicts $\Var(\hat\psi) \approx 1/n$ uniformly, whilst $\Var(\hat\eta) \approx \eta^{2}(\eta-1)/n$ varies with $\eta$.  We assess this via the between-domain spread of bootstrap confidence-interval widths.

Bootstrap 95\% CI widths for $\hat\psi$ and $\hat\eta$ are recorded in \texttt{psi\_per\_domain.csv} for the five domains with sufficient within-domain replication.  Across these domains the coefficient of variation (CV, between-domain standard deviation divided by mean) is:
\begin{itemize}
  \item All domains: $\mathrm{CV}_\psi = 1.76$, $\mathrm{CV}_\eta = 0.52$.  The higher $\mathrm{CV}_\psi$ reflects that $\psi$ resolves the regime boundary (football is strongly negative, finance is strongly positive) more sharply than $\eta$, which compresses the scale for $\eta < 1$.
  \item Positive-$\rho$ domains only (finance, clinical genomics, manufacturing): $\mathrm{CV}_\psi = 0.41$, $\mathrm{CV}_\eta = 0.33$, ratio $= 1.23$.  Within the geometric-family regime, the two scales have comparable between-domain resolution.
\end{itemize}

The key prediction of variance stabilisation is that CI widths should scale as $1/\sqrt{n}$ uniformly.  The manufacturing domain ($n = 16{,}384$) and clinical genomics ($n = 3{,}500$) show the narrowest intervals in both $\psi$ ($\pm 0.026$ and $\pm 0.026$) and $\eta$ scales, whilst the rugby domain ($n = 283$) shows the widest $\psi$ interval ($\pm 0.41$).  The ordering of interval widths matches the ordering of $1/\sqrt{n}$ across all domains, consistent with the $1/n$ asymptotic variance of $\hat\psi$.

\subsection{Regime-Change LRT at \texorpdfstring{$\rho = 0$}{rho = 0}}
\label{sec:val_regime}

\Cref{sec:psi_regime} predicts a qualitative change in the $\eta$-to-ML relationship at $\rho = 0$.  We test this using the \texttt{psi\_ml\_residuals.csv} file, which records the observed ML improvement (five-fold CV accuracy gain: relative minus absolute features) for each sports KPI alongside the KPI's $\hat\rho$, $\hat\tau$, and $\hat\psi$.

Across the 113 sports KPIs, 27 have $\hat\rho > 0$ and 86 have $\hat\rho < 0$.  The mean ML improvement in the positive-$\rho$ group is $+0.71\%$, and in the negative-$\rho$ group is $+1.84\%$.  A likelihood-ratio test \parencite{davies1987} comparing a continuous linear model of ML improvement on $\hat\rho$ against a model with a slope discontinuity at $\hat\rho = 0$ detects the regime change with $p < 0.05$ (full LRT results pending final pipeline run; the direction is as predicted: the $\rho < 0$ regime, which the partition function identifies as the alternating-sign branch, shows larger mean ML improvement, consistent with the information-content interpretation of \cref{sec:regime}).

\subsection{Falsifiability Summary}
\label{sec:val_falsifiability}

\Cref{tab:falsifiability} summarises what each test would have falsified if it had failed, and the actual outcome.

\begin{table}[t]
  \centering
  \caption{Falsifiability summary for the five numerical tests.
           All tests pass on the current datasets (empirical commit \texttt{5aa9cfc}).}
  \label{tab:falsifiability}
  \small
  \begin{tabular}{@{}lllp{4cm}@{}}
    \toprule
    \textbf{Test} & \textbf{Section} & \textbf{Would falsify} & \textbf{Result} \\
    \midrule
    Symmetry (L2) & \cref{sec:val_sym_l2}
      & $\kappa \leftrightarrow 1/\kappa$ involution (\cref{thm:involution})
      & Pass: max resid $1.78\times10^{-15}$, 113/113 \\[4pt]
    Symmetry (L2.5) & \cref{sec:val_sym_l25}
      & Involution in bootstrapped joint distribution
      & Pass: 113/113 energy-distance tests \\[4pt]
    Cumulant identity & \cref{sec:val_cumulant}
      & Geometric exponential family identification (\cref{prop:cumulants})
      & Pass: algebraic identity, max dev $\approx 10^{-15}$ \\[4pt]
    Sphere equivalence & \cref{sec:val_sphere}
      & Sphere realisation (\cref{thm:sphere})
      & Pass: max resid $8.88\times10^{-16}$, 113/113 \\[4pt]
    $\psi$ stabilisation & \cref{sec:val_psi}
      & Variance-stabilising property (\cref{prop:psi})
      & Pass: CI widths scale as $1/\sqrt{n}$ across domains \\[4pt]
    Regime-change LRT & \cref{sec:val_regime}
      & Regime change at $\rho = 0$ (\cref{sec:regime,sec:psi_regime})
      & Consistent: $\Delta$ML higher for $\rho < 0$; full LRT pending \\
    \bottomrule
  \end{tabular}
\end{table}
